Garbage Collection Evaluation Infrastructure for the Cog VM

### 2018-06-13 paper.tex
added todo for the other author
added ref and section bench

The Cog virtual machine (VM), the default VM for multiple programming languages such Pharo, Squeak and Newspeak, currently features in production a stop-the-world Mark-Compact algorithm as the full garbage collector (GC) algorithm. The algorithm has a high throughput but a high pause time during which the application is not responsive (the GC interrupts the application for multiple seconds on modern Macbooks starting from multiple Gbs heaps). For interactive applications, a new algorithm is required with a smaller pause time.

@@

\section{Building Macro-benchmarks}
\label{sec:bench}

2 directions. One is to build real benchs from real apps. Same direction as Dacapo... To be ported ?

@@

Explain all the phases

> Growing heap:
[GCMooseBench new benchLoadFromMSE: b wildfly].
benchLoadFromMSE: fileName
mooseModel := MooseModel new importFromMSEStream: (StandardFileStream readOnlyFileNamed: fileName)

[(GCMooseBench withMooseModel: model) benchExpandProperties]. "optional, increases drastically the model size"
benchExpandProperties
	allProperties :=  mooseModel entityStorage collect: [ :anEntity | 
    		anEntity mooseInterestingEntity complexPropertyPragmas collect: [ :aPragma |
        		[ (anEntity mooseInterestingEntity perform: aPragma methodSelector) ] on: Error do: ['error'] ] ].

> Stable heap:
[(GCMooseBench withMooseModel: model) benchAnalysisPyramid].
benchAnalysisPyramid
	mooseModel overviewPyramidMetrics.
[(GCMooseBench withMooseModel: model) benchAnalysisSymbol].
benchAnalysisSymbol
	mooseModel allMethods symbolsUsedInName
[(GCMooseBench withMooseModel: model) benchAnalysisDeprecation].
benchAnalysisDeprecation
	mooseModel allModelClasses select: [ :each |
    		(each isAnnotatedWith: 'Deprecated') and: [ each clientTypes notEmpty ] ].

> Shrinking heap:
[(GCMooseBench withMooseModel: model) benchRelease].
benchRelease
	mooseModel := nil.
	allProperties := nil.
	3 timesRepeat: [Smalltalk garbageCollect].

@@

Sophie's section

@@

Cog's Mark-Compact algorithm, named \emph{SpurSweeper}, is a sweep-only algorithm, setting the compactor to SpurSweeper effectively changes the fullGC to a mark-sweep non-moving algorithm.

SpurSweeper is a reference implementation if one wants to evaluate GC performance and compare it to a Mark-Sweep. It's also the only non-moving GC available right now which can be convenient for some experiments. One of the main reason why it was implemented is because advanced compaction algorithm includes a sweep phase (See SelectiveCompactor for example) and SpurSweeper allows to debug the sweep phase separately.

\subsection{Snapshot discussion}
Problem with the snapshot. and non compacting algo. Mention G1 and partial compaction too.

=> HybridCompactor solution.

@@

\paragraph{GC benchmark suite.}
Dacapo + see ref in Dacapo.

\paragraph{ACDC.}
ACDC-js (Sophie you write a paragraph on that.

>>>>

The Cog virtual machine (VM), the default VM for multiple programming languages such Pharo \cite{PharoByExample}, Squeak \cite{SqueakByExample} and Newspeak \cite{NewspeakOopsla}, currently features in production a stop-the-world Mark-Compact algorithm as the full garbage collector (GC) algorithm. The algorithm has a high throughput but a high pause time during which the application is not responsive (the GC interrupts the application for multiple seconds on modern Macbooks starting from multiple Gbs heaps). For interactive applications, a new algorithm is required with a smaller pause time.

@@

\section{Building Benchmarks}
\label{sec:bench}

To build GC benchmarks, we took two approaches. We asked companies for their deployed memory intensive applications and turned them into benchmarks. Multiple benchmarks in this category include closed-source code. In Section \ref{sec:mooseBench}, we describe how we turned a specific application into a benchmark with the help of the company (this specific benchmark does not include closed-source code). Then, to stress the GC on specific aspects, we built a configurable benchmark that based on the configuration emulates heaps with different properties.

@@

Part of the business of Feenk consists in analyzing softwares. To do so, the application parses a mse file into a model, performs some analysis, and then releases the model and generates analysis results. For business, the application analyzes closed-sourced application of companies. To build an open-source benchmark, we analyse open-source applications instead of closed-source ones.

We built six different benchmarks that falls into three categories: growing, stable and shrinking heaps. The benchmarks were built on top of the stable Moose image\footnote{http://www.moosetechnology.org/}.

\paragraph{Growing heap.} The growing heap benchmarks increase the heap size as they are performed, mainly creating objects. We built two benchmarks:
\begin{itemize}
	\item \emph{LoadFromMSE}: parses a mse file into a moose model, effectively loading a graph of object. This is the only benchmark not taking a moose model as a parameter.
	\item \emph{ExpandProperties}: computes all interesting properties of the model and store them, effectively loading a graph of object connected in many points to the original graph. 
\end{itemize}

\begin{code}
GCMooseBench >> benchLoadFromMSE: fileName
	mooseModel := MooseModel new importFromMSEStream: (StandardFileStream readOnlyFileNamed: fileName)

GCMooseBench >>benchExpandProperties
	allProperties :=  mooseModel entityStorage collect: [ :anEntity | 
    		anEntity mooseInterestingEntity complexPropertyPragmas collect: [ :aPragma |
        		[ (anEntity mooseInterestingEntity perform: aPragma methodSelector) ] on: Error do: ['error'] ] ].
\end{code}

\paragraph{Stable heap.} The stable heap benchmarks perform random memory accesses in a heap which size remains approximately the same. We built three benchmarks:

\begin{itemize}
	\item \emph{Pyramid}: computes an overview of the pyramid metrics of the software.
	\item \emph{Symbol}: computes all the symbols used in name. \todo{check}
	\item \emph{Deprecation}: computes all the deprecated code.
\end{itemize}

\begin{code}
benchAnalysisPyramid
	mooseModel overviewPyramidMetrics.

benchAnalysisSymbol
	mooseModel allMethods symbolsUsedInName

benchAnalysisDeprecation
	mooseModel allModelClasses select: [ :each |
    		(each isAnnotatedWith: 'Deprecated') and: [ each clientTypes notEmpty ] ].
\end{code}

\paragraph{Shrinking heap.} The shrinking heap benchmark frees part of the heap and performs a GC. We built only one benchmark, \emph{Release}, which removes references to the graph of objects and performs three garbage collections.

\begin{code}
benchRelease
	mooseModel := nil.
	allProperties := nil.
	3 timesRepeat: [Smalltalk garbageCollect].
\end{code}

@@

\todo{Sophie}
Sophie's section

@@

Cog's Mark-Compact algorithm, named \emph{SpurSweeper}, is a sweep-only algorithm, setting the compactor to SpurSweeper effectively changes the fullGC to a mark-sweep non-moving algorithm. It iterates over all old space in linear order. Each time an unmarked object or a free chunk is met, it coalesces it with subsequent ones and updates the free lists with a new larger free chunk.

\subsection{Snapshot discussion}

On top of the Cog VM, snapshots can be performed to persist the given state of the heap. Since snapshots are persisted, they should use a low amount of memory to avoid using too many bytes on disk. Non compacting algorithms, such as the Mark-Sweep, or algorithms compacting only part of the heap, such as garbage first~\cite{G1}, are a problem in this context. The heap, when collected by the Mark-Sweep, has a lot of free chunks when the heap size varies a lot leading to a large snapshot. However, non compacting have also interesting advantages (lower garbage collection pause time for example).

To be able to have a garbage collector with a low pause time and still be able to perform efficiently snapshots, we design a hybrid compactor solution. The VM tells the compactor if the garbage collection is performed for snapshot or not, and the compactor chooses to apply one algorithm or the other to have low compaction pause time or to snapshot with a low memory footprint.

@@

\paragraph{GC benchmark suite.}
Dacapo + see ref in Dacapo. \cite{DacapoBench}
\todo{Sophie}

\paragraph{ACDC.}
ACDC-js (Sophie you write a paragraph on that. \cite{ACDCJS}
